Attempting to predict Heart Disease using a classical decision tree machine learning method

### 2022-12-08 Manuscript/paper.tex
added actual citation names and did some light editing

\begin{document}
\maketitle
\begin{abstract}
  Heart disease has been the leading cause of death globally for 20 years \citep{}. The extraordinary availability of patient data in combination with the potential for great impact derived from successful statistical models serves as motivation for data analysis and model creation. The goal of this paper is to create a classical decision tree model using the largest heart disease data set available to predict heart disease. A classical decision tree with all variables in the data set had an error rate of .4967, a second model where individuals with a serum cholesterol level of 0 mg/dl were removed had an error rate of .5201, and a third model where the variable chest pain type was removed provided an error rate of .4924. The results provided by this paper reconfirm that classical decision trees are not the best model making method for heart disease. 
    \end{abstract}

@@

    To continue living a human’s heart needs to pump blood. Blood carries oxygen to our vital organs and also can take away the trash that cells produce as they do their daily tasks. Heart disease makes it hard for the human body to function properly. By researching and developing tools in which medical providers can input patient data and get outputs pertaining to if a patient has heart disease, medical providers can prescribe lifestyle changes to mitigate and decrease risk of significant cardiac events. 
1.72\% of the world’s population, which is about 126 million individuals, are affected by heart disease \citep{}. Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep{}. Heart disease is not only extraordinarily prevalent it is also a quite expensive disease to deal with. In 2010 it cost 863 billion us dollars on a global scale \citep{}. 
The prevalence of heart disease, in conjunction with the increased availability of anonymous patient data has led to many people attempting to create models which based on a set of input variables can predict heart disease. An example of such work is \citet{} where they compared performances of 3 different models including a decision tree, their tree model ended up having a .21 error rate. 
In this paper a multitude of classical decision trees were grown with the largest heart disease data set available to explore the error rate and if certain data and model manipulation could improve the error rate.

@@

PLEASE NOTE: STILL TRYING TO FIGURE OUT HOW TO OUTPUT THE GRAPHS IN ONE VISUAL SO THEY DONt TAKE UP PAGES AND PAGES OF SPACE.

@@

The aim for this paper is to create a classical decision tree machine learning model that given variable inputs can predict whether an individual has heart disease or not.
Three classical decision trees were created. One included all of the variables, one included all variables however the data had been modified to void any individuals that had a cholesterol level of 0 mg/dL. The final model created had been modified to include all variables except for chest pain type and ST slope.
Before any classical decision tree was made the data was split into 2 equal sized sets. One served as the training set and the other served as the validation or test set. A seed number of 4321 was used for the sample function in r so that results could be reproducible.
After the classical decision tree was made a determination if the tree needed to be trimmed by finding the smallest tree that had a cross validated error within one standard error of the minimum cross validated error.
To assess the model fit each model was validated with the validation set created at the beginning and the error rate was calculated by the sum of false negative and false positives over the total number of individuals in the set.

\section{Results}
\label{sec:resu}
As stated in the methods section, error rate was used to determine model fit and success.
The first model created, the one with all variables and all observations included produced an  error rate of .496732. This means that 49.67\% of the validation observations were misclassified by the model. After producing this model and investigating the cross validation table it was determined that no pruning was needed. Since the error rate was very large that motivated data modification and reassessment.
The next model that was made trimmed the number of observations by if an individual had a Cholesterol level of 0. This created a decision tree that needed to be pruned. Both the tree with 3 splits and 1 split met the requirements of falling within 1 standard error of the smallest cross validated error. The model was trimmed to have 3 splits because 1 split felt a little too simplistic. After pruning the tree the error rate was found to be .5201. This means that 52.01\% of the validation data was misclassified. This ended up being the worst performing model.
The final model created was modified to not include chest pain type because it is a variable that has 4 different classifications and usually variables that are numeric or can have binary classification work much better. Additionally ST slope was also removed because another variable in the data set provides a numeric measurement for this and the ST Slope variable, similar to the chest paint type variable, has 3 different classifications instead of 2. This tree also showed qualities which would make it a good candidate for pruning however, after attempting pruning in a bevvy of ways my program would seemingly not do any pruning. The error rate for this tree ended up being .4924. This means 49.24\% of the data in the validation set was misclassified. This model was surprisingly the one with the lowest error rate.

>>>>

\begin{document}
\maketitle
\begin{abstract}
  Heart disease has been the leading cause of death globally for 20 years \citep{khan2020global}. The extraordinary availability of patient data in combination with the potential for great impact derived from successful statistical models serves as motivation for data analysis and model creation. The goal of this paper is to create a classical decision tree model using the largest heart disease data set available to predict heart disease. A classical decision tree with all variables in the data set had an error rate of .4967, a second model where individuals with a serum cholesterol level of 0 mg/dl were removed had an error rate of .5201, and a third model where the variable chest pain type was removed provided an error rate of .4924. The results provided by this paper reconfirm that classical decision trees are not the best model making method for heart disease. 
    \end{abstract}

@@

    To continue living a human’s heart needs to pump blood. Blood carries oxygen to our vital organs and also can take away the trash that cells produce as they do their daily tasks. Heart disease makes it hard for the human body to function properly. By researching and developing tools in which medical providers can input patient data and get outputs pertaining to if a patient has heart disease, medical providers can prescribe lifestyle changes to mitigate and decrease risk of significant cardiac events. 
1.72\% of the world’s population, which is about 126 million individuals, are affected by heart disease \citep{khan2020global}. Additionally 9 million deaths were caused by heart disease globally which makes it the leading cause of death and it has retained this status for the last 20 years\citep{khan2020global}. Heart disease is not only extraordinarily prevalent it is also a quite expensive disease to deal with. In 2010 it cost 863 billion us dollars on a global scale \citep{khan2020global}. 
The prevalence of heart disease, in conjunction with the increased availability of anonymous patient data has led to many people attempting to create models which based on a set of input variables can predict heart disease. An example of such work is \citet{5643666} where they compared performances of 3 different models including a decision tree, their tree model ended up having a .21 error rate. 
In this paper a multitude of classical decision trees were grown with the largest heart disease data set available to explore the error rate and if certain data and model manipulation could improve the error rate.

@@

@@

The aim for this paper is to create a classical decision tree machine learning model that given variable inputs can predict whether an individual has heart disease or not.
Three classical decision trees were created. One included all of the variables and all individuals, one included all variables however the data has been modified to void any individuals that had a cholesterol level of 0 mg/dL. The final model created has been modified to include all variables except for chest pain type and ST slope.
Before any classical decision trees were made the data was split into 2 equal sized sets. One served as the training set and the other served as the validation set. A seed number of 4321 was used for the sample function in r so that results could be reproducible.
After the classical decision tree the decision regarding trimming the tree was made a by finding the smallest tree that had a cross validated error within one standard error of the minimum cross validated error.
To assess the model fit each model was validated with the validation set created at the beginning and the error rate was calculated by the sum of false negative and false positives over the total number of individuals in the set.

\section{Results}
\label{sec:resu}
As stated in the methods section, error rate was used to determine model fit and success.
The first model created, the one with all variables and all observations included produced an  error rate of .496732. This means that 49.67\% of the validation observations were misclassified by the model. After producing this model and investigating the cross validation table it was determined that no pruning was needed. Since the error rate was very large that motivated data modification and reassessment.
The next model that was made trimmed the number of observations by if an individual had a Cholesterol level of 0. This created a decision tree that needed to be pruned. Both the tree with 3 splits and 1 split met the requirements of falling within 1 standard error of the smallest cross validated error. The model was trimmed to have 3 splits because 1 split felt a little too simplistic. After pruning the tree the error rate was found to be .5201. This means that 52.01\% of the validation data was misclassified. This ended up being the worst performing model.
The final model created was modified to not include chest pain type because it is a variable that has 4 different classifications and usually variables that are numeric or have binary classification work much better. Additionally ST slope was also removed because another variable, Oldpeak, in the data set provides a numeric measurement for this. Also the ST Slope variable, similar to the chest pain type variable, has 3 different classifications instead of 2. This tree also showed qualities which would make it a good candidate for pruning however, after attempting pruning in a bevvy of ways my program would seemingly not do any pruning. The error rate for this tree ended up being .4924. This means 49.24\% of the data in the validation set was misclassified. This model was surprisingly the one with the lowest error rate.
